Hockey's Most Controversial Statistic: An Analysis of the Effectiveness of the Plus-Minus Statistic

### 2024-11-01 manuscript/statspaper.tex
added results

The main issue with the data is the violation of independence. This can be seen by creating a confusion matrix of the some 
of the more basic nhl statistics.

@@

There is a high correlation between different shot related statistics and point related statistics. For example, there is a 
0.826371 correlation between Goals and Shots. This makes since because in order to score a goal, the player needs to shoot first.
Correlation analysis can be used to identify the strength and direction of the relationship between various advanced metrics and 
the plus/minus statistic. This was done by calculating the correlation coefficient between +/- and each metric (like CF\%, GF\%, SCF\%, etc.).
The correlation values, ranging from -1 to 1, tell us how closely each metric aligns with the plus/minus. Higher positive correlations 
(e.g., with GF\% and PDO) indicate metrics that vary similarly to plus/minus, suggesting that they may capture overlapping aspects of 
offensive or defensive performance.

Due to the presence of correlation among variables, ridge regression was performed in order to identify how plus/minus can be employed 
to assess offensive and defensive contribution. It was also used to determine if other advanced metrics Corsi or Fenwick have the same, 
better, or worse predictive abilities compared to plus-minus. Ridge regression, a form of linear regression, is valuable when dealing 
with datasets with multicollinearity, where predictor variables have some sort of relationship with one another. This applies to hockey metrics 
like Corsi, Fenwick, scoring chances, etc. since these stats often interact or overlap in measuring aspects of performance.

Coupled with ridge regression, cross validation was utilized to assess the predictive power of different groups of variables (offensive, 
defensive, possession-based) on the plus/minus statistic. Cross-validation splits the data into training and testing sets multiple times, 
computing a model's predictive accuracy each time. In this case, the data was split into five subsets. In each iteration, one subset was
kept as the test set and the other four were used to train the ridge regression model. The R-sqared value was reported for each iteration
to measure how well offensive, defensive, and possession-based metrics (like Corsi and Fenwick) predict plus/minus.

In order to separate individual contributions to plus/minus from team-level effects, a mixed-effects model was used. Mixed-effects models 
allow for random effects, capturing variability at the team level, while fixed effects capture individual-level metrics. Individual metrics 
like CF\%, GF\%, and others were included that directly measure a player's on-ice performance. The model considered team-level averages 
(e.g., TeamMeanGF\%), accounting for how a player’s team context might influence their plus/minus.

@@

Correlation with Plus/Minus (+/-):

Metrics such as GF\%, SF\%, PDO, xGF\%, and HDGF\% show a relatively strong correlation with plus/minus, which aligns with 
expectations since these metrics also track goal-related contributions.

Cross-Validation Scores:

@@

@@

\begin{table}[tbp]
  \caption{Cross Validation Scores}
  \label{tab:rv}
\centering
\begin{tabular}{rrr}
  \toprule
Both & Offensive & Defensive \\ 
  \midrule
  0.56777061 & 0.64692431 & 0.73265605 \\ 
  0.51687324 & 0.62667921 & 0.70862251 \\ 
  0.49576518 & 0.62263538 & 0.69467537 \\ 
  0.49576518 & 0.58484799 & 0.65315835 \\ 
  0.61054685 & 0.69145036 & 0.76239753 \\ 
   \bottomrule
\end{tabular}
\end{table}

>>>>

\subsection{Correlation Analysis}
The main issue with the data is multicollinearity, where independent variables are highly correlated. A correlation matrix was created 
with some of the more basic nhl statistics to highlight this issue, shown in Table \ref{tab:correlation_matrix}.

@@

There is a high correlation between different shot related statistics and point related statistics. For example, there is a 
0.826371 correlation between Goals and Shots. This makes since because in order to score a goal, the player needs to shoot first.
Correlation analysis can be used to identify the strength and direction of the relationship between various advanced metrics and 
the plus/minus statistic. Metrics with high positive correlations (e.g., GF\% and PDO) suggest a strong alignment with the plus/minus, 
indicating that they may reflect similar aspects of offensive or defensive performance. This was done by calculating the correlation 
coefficient between +/- and each metric (like CF\%, GF\%, SCF\%, etc.).The correlation values, ranging from -1 to 1, tell us how closely 
each metric aligns with the plus/minus. Higher positive correlations (e.g., with GF\% and PDO) indicate metrics that vary similarly to 
plus/minus, suggesting that they may capture overlapping aspects of offensive or defensive performance.

\subsection{Ridge Regression}

Due to the presence of correlation among variables, ridge regression was performed in order to identify how plus/minus can be employed 
to assess offensive and defensive contribution. It was also used to determine if other advanced metrics Corsi or Fenwick have the same, 
better, or worse predictive abilities compared to plus-minus. Ridge regression, a form of regularized linear regression, is beneficial 
in handling datasets where predictor variables are highly interrelated, as is the case with advanced hockey metrics. Ridge regression 
estimates the contribution of metrics like Corsi, Fenwick, and scoring chances while controlling for their interdependencies. This 
applies to hockey metrics like Corsi, Fenwick, scoring chances, etc. since these statistics often interact or overlap in measuring aspects 
of performance.

\subsection{Cross-Validation}
Coupled with ridge regression, cross validation was utilized to assess the predictive power of different groups of variables (offensive, 
defensive, possession-based) on the plus/minus statistic. Cross-validation splits the data into training and testing sets multiple times, 
computing a model's predictive accuracy each time. In this case, the data was split into five subsets. In each iteration, one subset was
kept as the test set and the other four were used to train the ridge regression model. The R-sqared value was reported for each iteration
to measure how well offensive, defensive, and possession-based metrics (like Corsi and Fenwick) predict plus/minus. This process ensures 
that the model’s performance is consistent across different data partitions.

\subsection{Mixed Model Effects}
In order to separate individual contributions to plus/minus from team-level effects, a mixed-effects model was used. Mixed-effects models 
allow for random effects, capturing variability at the team level, while fixed effects capture individual-level metrics. Individual metrics 
like CF\%, GF\%, and others were included that directly measure a player's on-ice performance. The model considered team-level averages 
(e.g., TeamMeanGF\%), accounting for how a player’s team context might influence their plus/minus.

@@

The results of taking the correlation coefficients of plus-minus versus every other column is as follows:

\begin{table}[tbp]
  \caption{Cross Validation Scores}
  \label{tab:rv}
\centering
\begin{tabular}{rrr}
  \toprule
Variable & Correlation Coefficient \\ 
  \midrule
  GP & 0.232144
  G & 0.237058
  A & 0.342538
  P & 0.314842
  +/- & 1.000000
  P/GP & 0.277666
  EVG & 0.239076
  EVP & 0.336372
  PPG & 0.187284
  PPP & 0.226803
  SHG & 0.177479
  SHP & 0.282570
  OTG & 0.161693
  GWG & 0.305998
  S & 0.255841
  S\% & 0.048282
  TOI/GP & 0.253234
  TOI & 0.291711
  CF & 0.369200
  CA & 0.210752
  CF\% & 0.533927
  FF & 0.373707
  FA & 0.213121
  FF\% & 0.553993
  SF & 0.376144
  SA & 0.211659
  SF\% & 0.568873
  GF & 0.423772
  GA & 0.083042
  GF\% & 0.705010
  xGF & 0.386643
  xGA & 0.195763
  xGF\% & 0.585988
  SCF & 0.387204
  SCA & 0.191645
  SCF\% & 0.585252
  HDCF & 0.399467
  HDCA & 0.195659
  HDCF\% & 0.551392
  HDGF & 0.442448
  HDGA & 0.092987
  HDGF\% & 0.614221
  On-Ice SH\% & 0.324061
  On-Ice SV\% & 0.419340
  PDO & 0.546471
  Off. Zone Starts & 0.309539
  Neu. Zone Starts & 0.294922
  Def. Zone Starts & 0.236777
  On The Fly Starts & 0.273141
  Off. Zone Start \% & 0.110750
  Off. Zone Faceoffs & 0.353819
  Neu. Zone Faceoffs & 0.314974
  Def. Zone Faceoffs & 0.262650
  Off. Zone Faceoff \% & 0.182999
  \bottomrule
\end{tabular}
\end{table}

Metrics such as GF\%, FF\%, SF\%, PDO, xGF\%, and HDGF\% show a relatively strong correlation with plus/minus (0.705010, 0.553993, 0.568873, 
0.546471, 0.585988, 0.614221 respectively), which aligns with expectations since many of these metrics also track goal-related contributions. 
Higher positive correlations indicate metrics that vary similarly to plus/minus, suggesting that they may capture overlapping aspects of
offensive or defensive performance.

Ridge regression was performed on three different sets of variables: combined metrics, offensive only, and defensive only. The combined
metrics consisted of Goals For Percentage(GF\%), Scoring Chances Percentage (SF\%), High-Danger Chances while that player is on the ice (HDSC\%),
combined shooting and save percentage (PDO), Percent of shots that were goals for (On-Ice SH\%), High-Danger Goals for while that player is 
on the ice (HDGC\%). All of these metrics are similar to plus-minus in that they combine the defensive and offensive factors together.
The offensive only metrics were CF, FF, SF, GF, SCF, HDCF, HDGF, GF\%, G, A, P/GP, CF\%, SCF\%, and On-Ice SH\%. The defensive only metrics were
CA, FA, SA, GA, SCF, HDCA, HDGA, On-Ice SV\%. The results can be seen in the table below.

@@

\begin{table}[tbp]
  \caption{Cross Validation Scores}
  \label{tab:rv}
\centering
\begin{tabular}{rrr}
  \toprule
Both & Offensive & Defensive \\ 
  \midrule
  0.5696297 & 0.64692431 & 0.73265605 \\ 
  0.50125171 & 0.62667921 & 0.70862251 \\ 
  0.47617049 & 0.62263538 & 0.69467537 \\ 
  0.48944826 & 0.58484799 & 0.65315835 \\ 
  0.57586267 & 0.69145036 & 0.76239753 \\ 
  \bottomrule
\end{tabular}
\end{table}

@@

Plus-Minus vs Corsi vs Fenwick

Both Corsi and Fenwick were used as predictors in ridge regression with the variables GF\%, SCF\%, HDCF\%, PDO, On-Ice SH\%, On-Ice SV\%, G, A.
The results of the cross validation are shown in the table below

\begin{table}[tbp]
  \caption{Cross Validation Scores}
  \label{tab:rv}
\centering
\begin{tabular}{rrr}
  \toprule
Corsi & Fenwick \\ 
  \midrule
  0.93898296 & 0.95099819 \\ 
  0.90054897 & 0.93012992 \\ 
  0.93915158 & 0.9528202 \\ 
  0.92841222 & 0.95201592 \\ 
  0.9325304 & 0.95921829 \\ 
   \bottomrule
\end{tabular}
\end{table}

These values are notibly better that the cross validation scores for plus-minus, regardless of which set of predictors. This indicates
that both Corsi and Fenwick are significantly better at predicting player effectiveness than plus-minus.
